---
title: "External, cached, and remote data sources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{extdata_cachedata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  messages = FALSE,
  warning = FALSE
)
```

## Introduction

Not all of the data for mapbaltimore is loaded when you first load the package. The reason is that large spatial data sets can take a long time to load and slow down your environment after they are loaded into active memory. GitHub also has a maximum file size of 100 megabytes. While some data sets are rarely updated, others are regularly updated—making the data included with this package potentially outdated.

mapbaltimore works around these issues in a few different ways:

- Storing large (<100 MB) data sets as geopackage (.gpkg) files in the `inst/extdata` folder
- Caching very large (>100 MB) data sets as geopackage files in a cache folder selected using the `rappdirs::user_cache_dir()` function
- Offering other functions to download spatial data from remote sources including city and state ArcGIS servers, the state open data portal, and on Open Street Map

The external and cached data sources are never fully loaded into active memory. Instead, mapbaltimore uses the `wkt_filter` parameter for the `sf::st_read()` function as described in [this post on Working with Large Spatial Data in R](https://jayrobwilliams.com/posts/2020/09/spatial-sql) by Rob Williams to import only the data you need. Continue reading to explore how these three different data access options work within the package.

```{r setup}
library(mapbaltimore)
library(ggplot2)

set_map_theme()
```

## External data sources

Currently, the external data sources include:

- Trees (also available on the [Tree Baltimore website](https://baltimore.maps.arcgis.com/apps/webappviewer/index.html?id=d2cfbbe9a24b4d988de127852e6c26c8%20))
- Unimproved property (a subset of the cached real property data)
- Vegetated areas

This data can be accessed using the extdata parameter of the `get_area_data` function as the following examples illustrate.

```{r trees}
# Get trees in the Oliver neighborhood
oliver_trees <- get_area(type = "neighborhood", area_name = "Oliver") %>% 
get_area_data(extdata = "trees")

# Show the top 6 species pluralis (spp) by number
oliver_trees %>% 
  sf::st_drop_geometry() %>% 
  dplyr::count(spp, sort = TRUE, name = "count") %>% 
  dplyr::slice_head(n = 6) %>% 
  knitr::kable()

# Plot tree locations sized by diameter at breast height (dbh)
oliver_trees %>% 
  ggplot() +
  geom_sf(aes(size = dbh, color = condition), alpha = 0.5)
```

```{r vegetated_area}
druidhill <- get_area(type = "neighborhood", area_name = "Druid Hill Park")

get_area_data(area = druidhill, extdata = "vegetated_area") %>% 
  ggplot() +
  geom_sf(fill = "darkgreen", alpha = 0.5) +
  layer_area_mask(area = druidhill, alpha = 0.5, fill = "white", color = NA, dist = 5)
```

Incomplete documentation for these external datasets is included in the `extdata.R` file but this documentation not exported so cannot be accessed directly via the help menu.

## Cached data sources

After you first install mapbaltimore, you want to run `cache_mapbaltimore_data` with no parameters to access real property data, edge of pavement data, and streets outside Baltimore City. Customizing the cache location or the coordinate reference system is possible but it is not recommended. Downloading all of this data takes a *very* long time and takes up around around 700 MB when all three files are cached.

To update the cached files, use the `overwrite = TRUE` parameter. I recommend overwriting individual data sets rather than all cached data sets as the `real_property` data set is the only one that should require occasional updates.

```{r cache}
# cache_mapbaltimore_data()
# cache_real_property(overwrite = TRUE)
```

After the data is cached, it can always be accessed using the `cache_data` parameter of `get_area_data()`.

```{r edge_of_pavement}
get_area(type = "neighborhood", area_name = "Federal Hill") %>% 
  get_area_data(cachedata = "edge_of_pavement") %>% 
  ggplot() +
  geom_sf(color = "grey50")
```

The `map_area_property()` function requires the cached real property data and will not work without it.

```{r real_property}
get_area(type = "neighborhood", area_name = "Morgan Park") %>% 
map_area_property(property = "principal residence",
                  show_mask = TRUE)
```

Other functions such as `layer_area_streets()` use an optional parameter to determine whether to use the city streets data or the cached `baltimore_msa_streets` data. Maps that show area near or crossing the city boundaries show much better when the full data can be used.

```{r streets_msa}
second_district <- get_area(type = "congressional district", area_name = "2nd District")

district_map <- ggplot() +
  geom_sf(data = second_district, fill = "springgreen", alpha = 0.2)

# If msa = FALSE, layer_area_streets uses the city streets data
district_map + layer_area_streets(area = second_district, sha_class = c("INT", "FWY"))
  
# If msa = TRUE, layer_area_streets uses the cached baltimore_msa_streets data
district_map +
  layer_area_streets(area = second_district, sha_class = c("INT", "FWY"), msa = TRUE)
```

## Remote data sources

Both the externally stored and cached data could be access remotely whenever needed but the large file size makes keeping a local version much faster. If the data is more frequently updated (or less frequently used), downloading the data whenever it is needed is likely preferable.

For example, `get_area_esri_data()` can take a `url` for an ArcGIS MapServer or FeatureServer or take a `type` parameter that corresponds to a MapServer or FeatureServer url from Open Baltimore, Maryland iMap, or the [Maryland Food System Map](https://mdfoodsystemmap.org/). The list of urls is currently internal system data but will be opened to package users in the future. 

```{r esri_data}
north <- get_area(type = "planning district", area_id = "North")

ggplot() +
  layer_area_streets(area = north, sha_class = c("PART", "MART")) +
  geom_sf(data = get_area_esri_data(area = north, type = "red light cameras"),
          fill = "red", shape = 21, size = 4, color = "black") +
  geom_sf(data = get_area_esri_data(area = north, type = "fixed speed cameras"), fill = "yellow", shape = 25, size = 2, color = "black")
```

`get_area_osm_data` wraps the `osmdata` package largely to simplify the experience of using the package. Open Street Map is particularly useful for accessing data that may not be available from other non-commercial sources, such as the location of bowling alleys in Baltimore City and County. Be aware that the [Open Street Map license](https://wiki.osmfoundation.org/wiki/Licence/Licence_and_Legal_FAQ#What_is_the_licence.2C_how_can_I_use_it.3F) requires credit any time you use OSM data.  

```{r area_osm_data}
baltimore_county <- dplyr::filter(baltimore_msa_counties, namelsad == "Baltimore County")

baltimore_county %>% 
  get_area_osm_data(key = "leisure",
                value = "bowling_alley",
                trim = TRUE,
                crs = 2804) %>% 
  dplyr::filter(!is.na(name)) %>% 
  ggplot() +
  geom_sf(data = baltimore_msa_counties, color = "grey30", fill = NA) +
  ggrepel::geom_label_repel(aes(fill = name, label = name, geometry = geometry), stat = "sf_coordinates", force = 10, color = "white") +
    set_map_limits(area = baltimore_county) +
  guides(fill = "none") +
  labs(caption = "Data courtesy Open Street Map.")
```

Lastly, `get_maryland_open_resource()` can be used to access data from the Socrata-based Maryland Open Data portal by wrapping the `RSocrata::read.socrata()` function. The `get_area_crashes` function calls `get_maryland_open_resource()` using the appropriate identifier for the table on vehicular crashes.

```{r area_crashes}
charles_north <- get_area(type = "neighborhood", area_name = "Charles North")

charles_north %>% 
get_area_crashes(geometry = TRUE, trim = TRUE) %>% 
  ggplot() +
  layer_area_streets(area = charles_north) +
  geom_sf(color = "orangered", alpha = 0.75, size = 2.5) +
  geom_sf(data = charles_north, color = "black", fill = NA, size = 2, linetype = "dashed")
```

